---
title: "Liquid-Chromatography Retention Order Prediction for Metabolite Identification"
author: "Eric Bach, Sandor Szedmak, Celine Brouard, Sebastian Böcker and Juho Rousu"
date: "July 01, 2018"
output: html_document
---

```{r setup, include = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (ggplot2)
library (data.table)
library (knitr)
library (OrgMassSpecR)
library (shiny)

base_dir <- "~/Documents/studies/doctoral/projects/rt_prediction_ranksvm/method_publishing/"

source (paste (base_dir, "results/scripts/R/tools/evaluate_rank_correlations.R", sep = "/"))
source (paste (base_dir, "results/scripts/R/tools/kernel_tools.R", sep = "/"))
source (paste (base_dir, "results/scripts/R/helper.R", sep = "/"))

# Pre-processed retention time data
sdir <- paste (base_dir, "data/processed/PredRet/v2/", sep = "/")
# Results base-directoy
sdir_results <- paste (base_dir, "results/raw/PredRet/v2/final/", sep = "/")
sdir_results_base <- paste (base_dir, "results/raw/", sep = "/")
```

# Experimental results: Retention order prediction (Section 3.1)

```{r, echo = FALSE}
sysset <- 10
allpairsfortest <- "True"
```

## Accuracy of binary vs. counting fingerprints (Section 3.1.3)

Comparison of the pairwise prediction performance using binary and counting 
MACCS fingerprints as predictor (see Table 3). 

```{r, echo = FALSE}
# Load data for section 3.1.3
df_acc_3_1_3 <- data.frame()

for (predictor in c("maccs", "maccsCount_f2dcf0b3")) {
    tmp <- load_baseline_single_results (
        measure = c("accuracy", "accuracy_std"), 
        base_dir = paste0 (sdir_results, "ranksvm_slacktype=on_pairs/"),
        predictor = predictor, kernel = ifelse (predictor == "maccs", "tanimoto", "minmax"),  
        pair_params = list (allow_overlap = "True", d_lower = 0, d_upper = 16,
                            ireverse = "False", type = "order_graph"), 
        feature_type = "difference", flavor = list (
            allpairsfortest = allpairsfortest, featurescaler = "noscaling", sysset = sysset))
    
    tmp$predictor <- predictor
    tmp$sysset <- sysset
    
    df_acc_3_1_3 <- rbind (tmp, df_acc_3_1_3)
}
```

```{r, fig.align = "center", fig.width = 12, fig.height = 3, eval = FALSE, echo = FALSE}
ggplot (df_acc_3_1_3[d_lower == 0 & d_upper == Inf],
        aes (x = target, y = accuracy, fill = predictor, label = round (accuracy, 3))) + 
        geom_col (position = "dodge") +
        geom_errorbar (aes (ymin = accuracy - 2 * accuracy_std, 
                            ymax = accuracy + 2 * accuracy_std),
                       position = "dodge", alpha = 0.5) +
        geom_hline (yintercept = 0.5, linetype = 2, alpha = 0.5) +
        geom_text (size = 3, position = position_dodge (0.9), vjust = 4) + 
        coord_cartesian (ylim = c(0.45, 1)) +
        ylab ("Pairwise accuracy") + xlab ("Target system") +
        ggtitle ("Performance comparison of MACCS binary and counting fingerprints") +
        theme (plot.title = element_text (hjust = 0.5), 
               plot.subtitle = element_text (hjust = 0.5), 
               legend.position = "bottom")
```

```{r, echo = FALSE}
renderTable ({
    df_acc_3_1_3[d_lower == 0 & d_upper == Inf, 
                 .(target_system = target, pairwise_accuracy = round (accuracy, 3), 
                   pairwise_accuracy_2std = round (2 * accuracy_std, 3), predictor)]
}, digits = 3) 
```

## Retention order prediction: RankSVM vs. SVR (Section 3.1.4)

Comparison of RankSVM and SVR pairwise prediction accuracy.

### Single system for training 

A single system is used for the training. The performance is evaluated in target
system used for training. Remember, we use cross-validation for the performance 
evaluation. The results are shown in Table 4 in the paper (*Single system, target data only*).

```{r, echo = FALSE}
# Load data for section 3.1.4 (single system)
df_acc_svm <- load_baseline_results (
    measure = c("accuracy", "accuracy_std"), 
    base_dir = paste0 (sdir_results, "ranksvm_slacktype=on_pairs/"),
    predictor = "maccsCount_f2dcf0b3", kernel = "minmax",  
    pair_params = list (allow_overlap = "True", d_lower = 0, d_upper = 16,
                        ireverse = "False", type = "order_graph"), 
    feature_type = "difference", flavor = list (
        embso = "True", featurescaler = "noscaling", sysset = sysset))

df_acc_svm$sysset <- sysset
df_acc_svm$estimator <- "RankSVM"

df_acc_svr <- load_baseline_results (
    measure = c("accuracy", "accuracy_std"), 
    base_dir = paste0 (sdir_results, "svr/"),
    predictor = "maccsCount_f2dcf0b3", kernel = "minmax",  
    pair_params = NULL, feature_type = NULL, flavor = list (
        embso = "True", featurescaler = "noscaling", sysset = sysset))

df_acc_svr$sysset <- sysset
df_acc_svr$estimator <- "SVR"

df_acc_3_1_4_single_system <- rbind (df_acc_svm, df_acc_svr)

df_acc_svm <- NULL
df_acc_svr <- NULL
```

```{r, echo = FALSE}
renderTable ({
    df_acc_3_1_4_single_system[d_lower == 0 & d_upper == Inf & target == source,
                               .(target_system = target, pairwise_accuracy = round (accuracy, 3),
                                 pairwise_accuracy_2std = round (2 * accuracy_std, 3), estimator)]
}, digits = 3)
```

### Multiple systems for training

All systems in the dataset are used for training. We vary the percentage of target
system data used for training from 0% (10%) to 100%. The resulting curves are 
averaged across all target systems (Figure 4).

```{r, echo = FALSE, eval = TRUE}
# Load data
bl_rsvm_p <- data.table()
for (perc in seq (10, 100, by = 10)) {
    tmp <- load_baseline_single_perc_results (
        measure = c("accuracy", "accuracy_std"), 
        #measure = c("rank_corr"),
        base_dir = paste0 (sdir_results, "ranksvm_slacktype=on_pairs/"), 
        predictor = "maccsCount_f2dcf0b3", kernel = "minmax", pair_params = list (
            allow_overlap = "True", d_lower = 0, d_upper = 16, ireverse = "False", 
            type = "order_graph"),
        flavor = list (featurescaler = "noscaling", percfortrain = perc, sysset = "10"))    
    tmp$perc <- perc
    tmp$embso <- TRUE
    
    bl_rsvm_p <- rbind (bl_rsvm_p, tmp)
}
bl_rsvm_p$estimator <- "RankSVM"
bl_rsvm_p$scenario <- "Single system"
bl_rsvm_p$sysset <- "Single system"

bl_svr_p <- data.table()
for (perc in seq (10, 100, by = 10)) {
    tmp <- load_baseline_single_perc_results (
        measure = c("accuracy", "accuracy_std"), 
        #measure = c("rank_corr"),
        base_dir = paste0 (sdir_results, "svr/"), feature_type = NULL,
        predictor = "maccsCount_f2dcf0b3", kernel = "minmax", pair_params = NULL,
        flavor = list (featurescaler = "noscaling", percfortrain = perc, sysset = "10"))    
    tmp$perc <- perc
    tmp$embso <- TRUE
    
    bl_svr_p <- rbind (bl_svr_p, tmp)
}
bl_svr_p$estimator <- "SVR"
bl_svr_p$scenario <- "Single system"
bl_svr_p$sysset <- "Single system"

aoo_rsvm_p <- data.table()
for (perc in seq (0, 100, by = 10)) {
    for (embso in c("True", "False")) {
        tmp <- load_all_on_one_perc_results (
            measure = c("accuracy", "accuracy_std"), 
            #measure = c("rank_corr"), 
            base_dir = paste0 (sdir_results, "ranksvm_slacktype=on_pairs/"), 
            predictor = "maccsCount_f2dcf0b3", kernel = "minmax", pair_params = list (
                allow_overlap = "True", d_lower = 0, d_upper = 16, ireverse = "False", 
                type = "order_graph"),
            flavor = list (embso = embso, featurescaler = "noscaling", 
                           ltso = "False", percfortrain = perc, sysset = "10"))    
        tmp$perc <- perc
        tmp$embso <- as.logical (embso)
        
        aoo_rsvm_p <- rbind (aoo_rsvm_p, tmp)
    }
}
aoo_rsvm_p$estimator <- "RankSVM"
aoo_rsvm_p$scenario <- "Set of systems"
aoo_rsvm_p$sysset <- "Multiple systems"

aoo_svr_p <- data.table()
for (perc in seq (0, 100, by = 10)) {
    for (embso in c("True", "False")) {
        tmp <- load_all_on_one_perc_results (
            measure = c("accuracy", "accuracy_std"),
            #measure = c("rank_corr"),
            pair_params = NULL,
            base_dir = paste0 (sdir_results, "svr/"), feature_type = NULL, 
            predictor = "maccsCount_f2dcf0b3", kernel = "minmax", 
            flavor = list (embso = embso, featurescaler = "noscaling", 
                           ltso = "False", percfortrain = perc, sysset = "10"))    
        tmp$perc <- perc
        tmp$embso <- as.logical (embso)
        
        aoo_svr_p <- rbind (aoo_svr_p, tmp)
    }
}
aoo_svr_p$estimator <- "SVR"
aoo_svr_p$scenario <- "Set of systems"
aoo_svr_p$sysset <- "Multiple systems"
```

```{r, fig.align = "center", fig.width = 6, fig.height = 3, eval = TRUE, echo = FALSE}
df_acc_3_1_4_multiple_systems <- rbind (
    bl_rsvm_p[d_lower == 0 & d_upper == Inf],
    bl_svr_p[d_lower == 0 & d_upper == Inf],
    aoo_rsvm_p[d_lower == 0 & d_upper == Inf],
    aoo_svr_p[d_lower == 0 & d_upper == Inf])[embso == TRUE]
ggplot (df_acc_3_1_4_multiple_systems[, .(mean_accuracy = mean (accuracy)), 
                                      by = c("perc", "estimator", "scenario", "sysset")], 
        aes (x = perc, y = mean_accuracy, color = estimator, 
             linetype = sysset)) +
    geom_point (size = 2) + geom_line (size = 0.5) +
    xlab ("Percentage of target system molecules used for training") +
    ylab ("Pairwise accuracy\n(averaged across all target systems)") +
    guides (color = guide_legend (title = "Method"),
            linetype = guide_legend (title = "Training set")) +
    theme_bw()
```

Result shown in Table 4.
```{r, eval = TRUE, echo = FALSE}
renderTable({
    rbind (
        bl_rsvm_p[d_lower == 0 & d_upper == Inf & perc == 100,
              .(target_system = target, sysset, 
                pairwise_accuracy = round (accuracy, 3),
                pairwise_accuracy_2std = round (2 * accuracy_std, 3),
                estimator, perc = 100)],
        bl_svr_p[d_lower == 0 & d_upper == Inf & perc == 100,
              .(target_system = target, sysset, 
                pairwise_accuracy = round (accuracy, 3),
                pairwise_accuracy_2std = round (2 * accuracy_std, 3),
                estimator, perc = 100)],
        aoo_rsvm_p[d_lower == 0 & d_upper == Inf & perc == 100 & embso == TRUE,
              .(target_system = target, sysset, 
                pairwise_accuracy = round (accuracy, 3),
                pairwise_accuracy_2std = round (2 * accuracy_std, 3),
                estimator, perc = 100)],
        aoo_svr_p[d_lower == 0 & d_upper == Inf & perc == 100 & embso == TRUE,
              .(target_system = target, sysset, 
                pairwise_accuracy = round (accuracy, 3),
                pairwise_accuracy_2std = round (2 * accuracy_std, 3),
                estimator, perc = 100)],
        aoo_rsvm_p[d_lower == 0 & d_upper == Inf & perc == 0 & embso == TRUE,
              .(target_system = target, sysset, 
                pairwise_accuracy = round (accuracy, 3),
                pairwise_accuracy_2std = round (2 * accuracy_std, 3),
                estimator, perc = 0)],
        aoo_svr_p[d_lower == 0 & d_upper == Inf & perc == 0 & embso == TRUE,
              .(target_system = target, sysset, 
                pairwise_accuracy = round (accuracy, 3),
                pairwise_accuracy_2std = round (2 * accuracy_std, 3),
                estimator, perc = 0)])
}, digits = 3)
```

# Experimental results: Metabolite identification (Section 3.2)

```{r, echo = FALSE, eval = TRUE}
df_top1acc_FINAL_s10 <- NULL
df_top1acc_FINAL_s10_imp <- NULL
df_top1acc_FINAL_imp <- NULL
tmp <- NULL

D_space <- c("0", "1e-05", "0.0001", "0.0005", "0.001", "0.0025", "0.005", "0.0075",
             "0.01", "0.025", "0.05", "0.075", "0.1")

use_log <- "False"
n_rep <- 1000

df_top1acc_FINAL_s10_imp <- data.table()
for (D in D_space) {
    tmp <- load_topkacc_of_reranked_molecules_GS_BS (
        base_dir = paste0 (sdir_results_base, "/s10_imp_no3D/final/ranksvm_slacktype=on_pairs/"),
        predictor = "maccsCount_f2dcf0b3", kernel = "minmax", 
        predictor_column = NULL, kernel_column = NULL,
        pair_params = list (allow_overlap = "True", d_lower = 0, d_upper = 16,
                            ireverse = "False", type = "order_graph"), 
        feature_type = "difference", flavor = list (
            D = D, epsrt = "0", featscaler = "noscaling", nrds = as.character(n_rep), sysset = "10_imp",
            usecoldesc = "False", uselog = use_log, wfun = "pwmax"))
     
    tmp$D <- as.numeric (D)
    tmp$wfun <- "pwmax"
    tmp$sysset <- "Others & target"
    
    df_top1acc_FINAL_s10_imp <- rbind (df_top1acc_FINAL_s10_imp, tmp)
}

df_top1acc_FINAL_imp <- data.table()
for (D in D_space) {
    tmp <- load_topkacc_of_reranked_molecules_GS_BS (
        base_dir = paste0 (sdir_results_base, "/s10_imp_no3D/final/ranksvm_slacktype=on_pairs/"),
        predictor = "maccsCount_f2dcf0b3", kernel = "minmax", 
        predictor_column = NULL, kernel_column = NULL,
        pair_params = list (allow_overlap = "True", d_lower = 0, d_upper = 16,
                            ireverse = "False", type = "order_graph"), 
        feature_type = "difference", flavor = list (
            D = D, epsrt = "0", featscaler = "noscaling", nrds = as.character(n_rep), sysset = "imp",
            usecoldesc = "False", uselog = use_log, wfun = "pwmax"))
     
    tmp$D <- as.numeric (D)
    tmp$wfun <- "pwmax"
    tmp$sysset <- "Target"
    
    df_top1acc_FINAL_imp <- rbind (df_top1acc_FINAL_imp, tmp)
}

df_top1acc_FINAL_s10 <- data.table()
for (D in D_space) {
    tmp <- load_topkacc_of_reranked_molecules_GS_BS (
        base_dir = paste0 (sdir_results_base, "/PredRet/v2/final/ranksvm_slacktype=on_pairs/"),
        predictor = "maccsCount_f2dcf0b3", kernel = "minmax", 
        predictor_column = NULL, kernel_column = NULL,
        pair_params = list (allow_overlap = "True", d_lower = 0, d_upper = 16,
                            ireverse = "False", type = "order_graph"), 
        feature_type = "difference", flavor = list (
            D = D, epsrt = "0", featscaler = "noscaling", nrds = as.character(n_rep), sysset = "10",
            usecoldesc = "False", uselog = use_log, wfun = "pwmax"))
     
    tmp$D <- as.numeric (D)
    tmp$wfun <- "pwmax"
    tmp$sysset <- "Others"
    
    df_top1acc_FINAL_s10 <- rbind (df_top1acc_FINAL_s10, tmp)
}

```

Average percentage of correctly identified molecular structures for different values
of $D$ (Figure 5).

```{r, fig.align = "center", fig.width = 6, fig.height = 3, eval = TRUE, echo = FALSE}
data_greater0 <- NULL
data_equal0 <- NULL

data_greater0 <- rbind (
    df_top1acc_FINAL_s10_imp, df_top1acc_FINAL_s10, df_top1acc_FINAL_imp
    )[D > 0.0001 & D < 0.25, 
      .(top1_acc_mean = mean (top1_acc), top1_acc_sde = std_err (top1_acc)),
      by = c("sysset", "D", "wfun", "n_spectra")]

data_equal0 <- rbind (
    df_top1acc_FINAL_s10_imp, df_top1acc_FINAL_s10, df_top1acc_FINAL_imp
    )[D == 0, 
      .(top1_acc_mean = mean (top1_acc), top1_acc_sde = std_err (top1_acc)),
      by = c("sysset", "D", "wfun", "n_spectra")]
data_equal0 <- unique (data_equal0)

ggplot (data_greater0, 
        aes (x = D, y = top1_acc_mean, color = sysset, fill = sysset)) +
    geom_hline (data = data_equal0, aes (yintercept = top1_acc_mean), lty = 1, alpha = 1) +
    geom_point () +  geom_line (lty = 2, alpha = 1) +
    scale_x_log10 (breaks = as.numeric (D_space[2:length(D_space)])) +
    xlab ("D: Weighting of the order information") +
    ylab ("Percentage of correctly\nidentified compounds") +
    scale_fill_discrete(guide = FALSE) +
    theme_bw() +
    theme (axis.text.x = element_text (angle = 45, hjust = 1),
           legend.position = "right") + 
    guides (color = guide_legend (title = "RankSVM\ntraining set"))
```

Significants test for $D$ with maximum identification accuracy:

- Others \& target:
```{r, eval = TRUE}
t.test(df_top1acc_FINAL_s10_imp[D == 0.0075]$top1_acc, mu = mean (df_top1acc_FINAL_s10_imp[D == 0]$top1_acc))
```
- Others:
```{r, eval = TRUE}
t.test(df_top1acc_FINAL_s10[D == 0.01]$top1_acc, mu = mean (df_top1acc_FINAL_s10[D == 0]$top1_acc))
```
