---
title: "Pre-processing: PredRet v2"
author: "Eric Bach"
date: "December 16, 2017"
output: html_document
runtime: shiny
---

```{r}
####
#
# The MIT License (MIT)
#
# Copyright 2017 Eric Bach <eric.bach@aalto.fi>
#
# Permission is hereby granted, free of charge, to any person obtaining a copy 
# of this software and associated documentation files (the "Software"), to deal 
# in the Software without restriction, including without limitation the rights 
# to use, copy, modify, merge, publish, distribute, sublicense, and/or sell 
# copies of the Software, and to permit persons to whom the Software is furnished
# to do so, subject to the following conditions:
#    
# The above copyright notice and this permission notice shall be included in all
# copies or substantial portions of the Software.
#
# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND, EXPRESS OR 
# IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF MERCHANTABILITY, 
# FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT. IN NO EVENT SHALL THE 
# AUTHORS OR COPYRIGHT HOLDERS BE LIABLE FOR ANY CLAIM, DAMAGES OR OTHER 
# LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, 
# OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE
# SOFTWARE.
#
####
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library (ggplot2)
library (data.table)
library (shiny)
```

## PredRet v2: Pre-processing of the RT data

Script to pre-process the Aug16 PredRet data:

* Consider only reversed phase columns
* Consider only datasets included in the publication
* Early eluting molecules are excluded
* Minimum retention time rather than average for multiple RTs
* Consistency check for the itendification, e.g. RT difference to large

### Read and plot the RT data
```{r}
base_dir <- stop ("DEFINE BASE-PATH CONTAINING INPUT DATA!") 
# Eg: base_dir <- "~/Documents/studies/doctoral/projects/rt_prediction_ranksvm/data"
raw_data_dir <- paste (base_dir, "/raw/PredRet/", sep = "/")

db <- data.table (read.csv (
    paste (raw_data_dir, "Stanstrup_exp_Aug16.csv", sep = "/"),
    sep = ";"))
setkey (db, "system")

# Consider only reversed phase columns
reversed <- c("LIFE_old", "LIFE_new", "FEM_long", "FEM_short",
              "FEM_orbitrap_plasma", "FEM_orbitrap_urine", "FEM_lipids",
              "IPB_Halle", "RIKEN", "Eawag_XBridgeC18", "MPI_Symmetry",
              "MTBLS4", "MTBLS17", "MTBLS19", "MTBLS20", "MTBLS36",
              "MTBLS38", "MTBLS39", "UFZ_Phenomenex", "UniToyama_Atlantis",
              "MTBLS52")
db <- db[reversed]

# Consider only datasets included in the publication
db <- db[as.POSIXlt.date(db$date.added) <= as.POSIXlt.date("2015-06-18"), ]

# Exclude measurements which are flagged as suspected
db <- db[suspect == FALSE]
```

```{r, echo=FALSE, fig.width=9, fig.align="center", eval=FALSE}
ggplot (db, aes (x = recorded_rt)) + 
    geom_histogram (binwidth = 1) + 
    facet_wrap (~ system, scales = "free_y") +
    xlab ("Retention time") + ylab ("Number of measurements")
```
The datasets (partially representing different chromatographic columns) cover 
very different retention time ranges, i.e. run-time of the columns. For each 
dataset a seperate RT threshold can be defined to exclude early eluting molecules. 

```{r, echo=FALSE, fig.width=9, fig.align="center", eval=FALSE}
ggplot (db, aes (x = recorded_rt)) + 
    geom_histogram (bins = 30) + 
    facet_wrap (~ system, scales = "free") +
    xlab ("Retention time") + ylab ("Number of measurements")
```

### Remove early eluting molecules

The removal of early eluting molecules is currently done individually for each 
dataset and performed manually. Large peaks in the begining, i.e. at low RTs, 
correspond to the molecules that do not interact with the chromatographic system.

We therefore inspect each dataset separetly at the beginning:

```{r, eval=TRUE, echo=FALSE}
selectInput ("dataset", label = "Dataset:", choices = reversed)

renderPlot ({ 
    ggplot (db[input$dataset], aes (x = recorded_rt)) + 
        geom_histogram (bins = 70) +
        scale_x_continuous (breaks = scales::pretty_breaks (n = 30))
})
```

```{r, eval=TRUE, echo=FALSE}
# db_rtthr <- data.table (system = reversed, rtthr = 0.0)
# setkey (db_rtthr, "system")

setkey (db, "system")
db$rtthr <- 0.0
db["LIFE_old"]$rtthr <- 0.6
db["LIFE_new"]$rtthr <- 0.6
db["FEM_long"]$rtthr <- 3.0
db["FEM_short"]$rtthr <- 0.0
db["FEM_orbitrap_plasma"]$rtthr <- 2.0
db["FEM_orbitrap_urine"]$rtthr <- 0.0
db["FEM_lipids"]$rtthr <- 1.0
db["IPB_Halle"]$rtthr <- 0.9
db["RIKEN"]$rtthr <- 0.3
db["Eawag_XBridgeC18"]$rtthr <- 1.5
db["MPI_Symmetry"]$rtthr <- 1.5
db["MTBLS4"]$rtthr <- 0.0
db["MTBLS17"]$rtthr <- 0.5
db["MTBLS19"]$rtthr <- 1.0
db["MTBLS20"]$rtthr <- 1.25
db["MTBLS36"]$rtthr <- 1.75
db["MTBLS38"]$rtthr <- 1.75
db["MTBLS39"]$rtthr <- 2.5
db["UFZ_Phenomenex"]$rtthr <- 2.5
db["UniToyama_Atlantis"]$rtthr <- 0.0
db["MTBLS52"]$rtthr <- 0.0

# db_rtmax <- db[, .(rt_max = max (recorded_rt)), by = "system"]
# setkey (db_rtmax, "system")

# stopifnot (all (db_rtmax$system == db_rtthr$system))

# db_rtthr$rtthr_p <- db_rtthr$rtthr / db_rtmax$rt_max * 100

renderTable ({
    db[, .(rtthr = unique (rtthr)), by = "system"]
}) 
```

```{r,eval=TRUE}
# Early eluting molecules are excluded
db_early <- db[recorded_rt >= rtthr]
```

```{r, eval=TRUE, echo=FALSE}
selectInput ("dataset", label = "Dataset:", choices = reversed)

renderPlot ({ 
    ggplot (db_early[input$dataset], aes (x = recorded_rt)) + 
        geom_histogram (bins = 70) +
        scale_x_continuous (breaks = scales::pretty_breaks (n = 30))
})
```

### Aggregating multiple RT measurements

```{r, eval = TRUE}
# Minimum retention time rather than average for multiple RTs
db_agg <- db_early[, 
    .(rt        = min (recorded_rt), 
      rt_diff   = max (recorded_rt) - min (recorded_rt), 
      rt_diff_p = (max (recorded_rt) - min (recorded_rt)) / 
          min (recorded_rt) * 100,
      n_rep     = .N), 
    by = c("inchi", "system")]
```

Number of molecules per system after aggregation
```{r, eval=TRUE, echo=FALSE}
renderTable({db_agg[, .N, by = c("system")]})
```

### Remove incosistend identifications
```{r, eval=TRUE}
# Consistency check for the itendification, e.g. RT difference to large
db_agg_cons <- db_agg[n_rep == 1 | rt_diff_p <= 5]
```

```{r, echo=FALSE}
renderTable({db_agg_cons[, .N, by = c("system")]})
```

```{r, echo=FALSE, fig.width=9, fig.align="center", eval=FALSE}
ggplot (db_agg_cons, aes (x = rt)) + 
    geom_histogram (bins = 30) + 
    facet_wrap (~ system, scales = "free") +
    xlab ("Retention time") + ylab ("Number of measurements")
```

## Write out pre-processed dataset
```{r, echo=TRUE, eval=FALSE}
out_dir <- paste (base_dir, "/processed/PredRet/v2/", sep = "/")
write.csv (db_agg_cons[, .(inchi, rt, system)],
           file = paste (out_dir, "rts.csv", sep = "/"),
           row.names = FALSE, quote = TRUE)
write.table (db_agg_cons[, .N, by = c("system")],
             file = paste (out_dir, "mol_per_system.csv", sep = "/"),
             row.names = FALSE, quote = FALSE, sep = "\t")
```
